How to select predictive models for decision-making or causal inference?

### 2024-10-20 response_to_reviewer.md

We thank the reviewers for their time, thoughts, and feedback on the
manuscript. We have addresed the comments through additional work as well
as clarifications in the revised manuscript.

To clarify the origin of the Robinson decomposition, as suggested by
reviewer 1, we have added an equation. As a result the equation numbers
in the revised manuscript are all shifted by one.

@@

# Reviewer 1

*[...] The authors conceivably state that "The problem is that causal estimation requires controlling an error on both treated and non-treated outcome for the same individual.". Consequently, they report on estimator that attempt to consider this (in Section "Methods"). However, I cannot understand how this hint ("controlling error on both treated and non-treated sample (or patient?)") has been implemented by the procedure the authors describe in Equations (1), (2), and (3). If, I correctly understand", in Eq. (2) we see the g-computation estimators, a well-known concept introduced by [21] (and recently used also by https://www.nature.com/articles/s41598-020-65917-x ), that splits the two configurations "treated" and "non-treated". [...] I understand that Eq. (2) and in Eq. (3) consider both treated and non-treated sample (or patient?), but those sums can be also equal to 0, even if their single terms are different from 0. Moreover, it is not clear if, and how Eq. (3) is related to Eq. (2): is it a variant? Is Eq. (3) derived from Eq. (2) under special assumptions?*

> The reviewer is right that Eq. (2) is part of constructing the classic
> g-computation estimator, as well as Eq. (1) and Eq. (3). We added a
> sentence at the begining of the section to stress that it is a section
> summarizing a classic framework.

> Eq. (3) (now 4) is an expression of the estimator of the ATE based on a model of the outcome, sometimes called g-estimation. It is unbiased under consistency (Assumption. 3) and unconfoundedness (Assumption 1). To prove that, take the expectation of Eq. (3).  Then apply  Eq. (1) (now 2) which is true under consistency (Eq. 1 and Assumption 1) and unconfoundedness (Assumption 3). Then by the law of total expectancy, the conditionnality on X disappear and we recover Eq. (2) (now 3)

> G-estimation itself does not control error on both populations. The formulas in Eq. (1-3) (now 2-4) are all partly oracle formulas: they rely on conditional expectations but give no specific procedures on how to compute them. We added a clarification on this topic above Eq. (4) (now 5) and above Eq. (8) (now 9).

### 2024-10-20 how_to_select_latex.tex
iter

A given data needs to meet a few assumptions to enable identifying
causal estimands \cite{rubin_causal_2005}. The usual
\emph{strong ignorability} assumptions are  (details in
\ref{apd:causal_assumptions}): \emph{1)}
\emph{unconfoundedness} \mbox{$\{Y(0),
        Y(1) \} \indep A | X$}, \emph{~2)} \emph{strong overlap} ie. every patient has a
strictly positive probability to receive each treatment, \emph{3)}
\emph{consistency}, and \emph{4)} \emph{generalization}.

@@

Given a sample of data and the oracle response functions $\mu_0, \mu_1$, the
finite sum version of \autoref{eq:tau_population} leads to an unbiased
estimator of the ATE written:
\begin{equation}
    \hat \tau = \frac{1}{n} \biggl(\sum_{i=1}^n \mu_{1}(x_i) - \mu_{0}(x_i) \biggr)
    \label{eq:ate_estimate}
\end{equation}
This estimator is an oracle \emph{finite sum} estimator by opposition to the
population expression of $\tau$, $\mathbb{E}[\mu_{1}(x_i) - \mu_{0}(x_i)]
$,
which involves an expectation taken on the full
distribution $\mathcal D$, which is observable but requires infinite data. For
each estimator $\ell$ taking an expectation over $\mathcal D$, we use the symbol
$\hat \ell$ to note its finite sum version. 
The formulas in Eq. (1-3) are all partly oracle formulas: they rely on
conditional expectations, the response functions, but give no specific
procedures on how to compute or select them. This last point is the topic
of our work, describe in the next section.
%In practice the response function can be estimated with one statistical model taking the treatment as a supplementary covariate $\hat \mu(x, a)$ (called S-learner) or by two separate model fitted on each population, treated or not $(\hat \mu_0(x), \hat \mu_1(x))$ (T-learner) \cite{kunzel_metalearners_2019}.

@@

\paragraph{Robinson decomposition}
G-computation is a choice of decomposition of the CATE estimation. Other choices
of decomposition exist, such as the R-decomposition
\cite{robinson_rootnconsistent_1988}. The latter introduces two new statistical
estimates, the conditional mean outcome and the probability of being treated (known
as propensity score \cite{rosenbaum_central_1983}):
\begin{flalign}
    \text{Conditional mean outcome} &                    & m(x) \myeq \; & \mathbb E_{Y \sim
    \mathcal{D}} [Y|X=x]            &                    &
    \label{def:m}
    \\
    \text{Propensity score}         &                    &
    e(x) \myeq \;                   & \mathbb P[A=1|X=x]
    \label{def:propensity_score}
\end{flalign}
the outcome can be written
\begin{flalign}\label{eq:r_decomposition}
    \text{R-decomposition}            &  & y(a) = m(x) + \big( a - e(x) \big)
    \tau(x) + \varepsilon(x; a)\notag &  &
    \\\ &&\text{with}\quad \mathbb E[\varepsilon(X; A)|X, A] = 0&&
\end{flalign}
$m$ and $e$ are often called
\emph{nuisances} \cite{chernozhukov_double_2018}. They are unknown and are estimated in practice thanks to statistical models (eg. linear models or forest based estimators).

@@

We assume the following four assumptions, referred as strong ignorability and
necessary to assure identifiability of the causal estimands with observational
data \cite{rubin_causal_2005}:
\begin{assumption}[Unconfoundedness]\label{assumption:ignorability}
    \begin{equation*}\label{eq:ignorability}
        \{Y(0), Y(1) \} \indep A | X
    \end{equation*}
    This condition --also called ignorability-- is equivalent to the conditional
    independence on $e(X)$ \cite{rosenbaum_central_1983}: $\{Y(0), Y(1) \}
        \indep  A | e(X)$.
\end{assumption}

@@

\begin{assumption}[Consistency]\label{assumption:consistency} The observed
    outcome is the potential outcome of the assigned treatment:
    \begin{equation*}\label{eq:consistancy}
        Y = A \, Y(1) + (1-A) \, Y(0)
    \end{equation*}
    Here, we assume that the intervention $A$ has been well defined. This
    assumption focuses on the design of the experiment. It clearly states the link
    between the observed outcome and the potential outcomes through the
    intervention \cite{hernan_causal_2020}.
\end{assumption}

>>>>

A given data needs to meet a few assumptions to enable identifying
causal estimands \cite{rubin_causal_2005}. \emph{1)} an individual's
outcome $Y$ is solely governed by the corresponding potential outcome:
\begin{flalign}
    \text{\emph{Consistency} assumption,}
    &&
    Y = A\, Y(1) + (1 - A)\, Y(0)
    \label{eq:consistency}
    &&
\end{flalign}
\emph{2)}
\emph{unconfoundedness} \mbox{$\{Y(0),
        Y(1) \} \indep A | X$}, \emph{~3)} \emph{strong overlap} ie. every patient has a
strictly positive probability to receive each treatment, 
and \emph{4)} \emph{generalization} --no covariate shift.
These classic assumptions, called
\emph{strong ignorability}, are formally detailed in
\ref{apd:causal_assumptions}.

@@

Given a sample of data and the oracle response functions $\mu_0, \mu_1$, the
finite sum version of \autoref{eq:tau_population} leads to an unbiased
estimator of the ATE written:
\begin{equation}
    \hat \tau = \frac{1}{n} \biggl(\sum_{i=1}^n \mu_{1}(x_i) - \mu_{0}(x_i) \biggr)
    \label{eq:ate_estimate}
\end{equation}
This estimator is an oracle \emph{finite sum} estimator by opposition to the
population expression of $\tau$, $\mathbb{E}[\mu_{1}(x_i) - \mu_{0}(x_i)]
$,
which involves an expectation taken on the full
distribution $\mathcal D$, which is observable but requires infinite data. For
each estimator $\ell$ taking an expectation over $\mathcal D$, we use the symbol
$\hat \ell$ to note its finite sum version. 
The formulas in Eq. (\ref{eq:mu_identification}-\ref{eq:ate_estimate}) are all partly oracle formulas: they rely on
conditional expectations, the response functions, but give no specific
procedures on how to compute or select them. This last point is the topic
of our work, describe in the next section.
%In practice the response function can be estimated with one statistical model taking the treatment as a supplementary covariate $\hat \mu(x, a)$ (called S-learner) or by two separate model fitted on each population, treated or not $(\hat \mu_0(x), \hat \mu_1(x))$ (T-learner) \cite{kunzel_metalearners_2019}.

@@

\paragraph{Robinson decomposition}
G-computation is a choice of decomposition of the CATE estimation. Other choices
of decomposition exist, such as the R-decomposition
\cite{robinson_rootnconsistent_1988}. The latter introduces two new statistical
estimates, the conditional mean outcome and the probability of being treated (known
as propensity score \cite{rosenbaum_central_1983}):
\begin{flalign}
    \text{Conditional mean outcome} &                    & m(x) \myeq \; & \mathbb E_{Y \sim
    \mathcal{D}} [Y|X=x]            &                    &
    \label{def:m}
    \\
    \text{Propensity score}         &                    &
    e(x) \myeq \;                   & \mathbb P[A=1|X=x]
    \label{def:propensity_score}
\end{flalign}
with these, the outcome (Eq.\,\ref{eq:consistency}) can be written
\begin{flalign}\label{eq:r_decomposition}
    \text{R-decomposition}            &  & y(a) = m(x) + \big( a - e(x) \big)
    \tau(x) + \varepsilon(x; a)\notag &  &
    \\\ &&\text{with}\quad \mathbb E[\varepsilon(X; A)|X, A] = 0&&
\end{flalign}
$m$ and $e$ are often called
\emph{nuisances} \cite{chernozhukov_double_2018}. They are unknown and are estimated in practice thanks to statistical models (eg. linear models or forest based estimators).

@@

We assume the following four assumptions, referred as strong ignorability and
necessary to assure identifiability of the causal estimands with observational
data \cite{rubin_causal_2005}:
\begin{assumption}[Consistency]\label{assumption:consistency} The observed
    outcome is the potential outcome of the assigned treatment:
    \begin{equation*}\label{eq:consistancy}
        Y = A \, Y(1) + (1-A) \, Y(0)
    \end{equation*}
    Here, we assume that the intervention $A$ has been well defined. This
    assumption focuses on the design of the experiment. It clearly states the link
    between the observed outcome and the potential outcomes through the
    intervention \cite{hernan_causal_2020}.
\end{assumption}

\begin{assumption}[Unconfoundedness]\label{assumption:ignorability}
    \begin{equation*}\label{eq:ignorability}
        \{Y(0), Y(1) \} \indep A | X
    \end{equation*}
    This condition --also called ignorability-- is equivalent to the conditional
    independence on $e(X)$ \cite{rosenbaum_central_1983}: $\{Y(0), Y(1) \}
        \indep  A | e(X)$.
\end{assumption}

@@
